---
title: "Sepsis cluster project"
output: html_document
written by: "Kristin Wickstrøm" 
date: "2024-12-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r import packages}
install.packages(c("mice","dplyr","tidyverse","readxl"))
#if (!require("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("ConsensusClusterPlus")
library(mice)
library(ConsensusClusterPlus)
library(dplyr)
library(tidyverse)
library(readxl)
```

```{r import dataset}
#import dataset that has 27 variables compared to Seymour where the nonnormal variables are logtransformed (creatinine, lactate, bilirubin, crp, alp, alat, troponinT, INR, leucocytes, urea, thrombocytes, paO2, glucose) and exp.transformed (saO2).
sepsis <- "import data"
```

#Multiple imputation and Consensus cluster 
The code under will run multiple imputation, 100 datasets with the method pmm. Max iterations= 20. Further it will run Consensus cluster on each of the imputed datasets and produce a matrix as a results. Each of the matrices will be combined to a mean matrix= mlC. 

```{r multiple imputation}

#########IMPUTATION--------------------------------
  
impute <- TRUE

if(impute){
  
  #100 imputations
  imputatedsepsis1 = mice(sepsis, m= 100, method=c("pmm"), maxit=20)
  
  
  #make 100 complete dataset
  sepsisimp <- list()
  for(i in 1:100) {
    sepsisimp[[i]] <- complete(imputatedsepsis1,i)
  }
  
  ###PREPROCESSING CONSENSUS CLUSTER PLUS-------
  
  sepCC <- list()
  sepCCm <- list()
  CCresultatliste1 <- list()
  sepsis_scaled <- list()
  
  for(i in 1:100) {
    sepsis_scaled[[i]] <- scale(sepsisimp[[i]])
    sepCC[[i]] <- as.data.frame(t(sepsis_scaled[[i]])) 
    sepCCm[[i]] <- as.matrix(sepCC[[i]]) 
  }
  
  ######## SAVE IMPUTED DATA ###########
  save(sepsis_scaled, sepCC, sepCCm, CCresultatliste1, file = 'imputed_data26.01.RData')
  
}else{
  
  load('imputed_data.RData')
  
}
  
####ConsensusClusterPlus, on 100 datasets, 10 clusters
for(i in 1:100) {    
  CCresultatliste1[[i]] = ConsensusClusterPlus(sepCCm[[i]], maxK=10, reps=100, pItem=0.8, clusterAlg="km", distance="euclidean", seed=10000, plot=NULL) 
}    

matrix_liste <- vector("list", 9)  # Create a list to hold 9 lists for k = 2 to 10

# Collect consensus matrices for each k
for (k in 2:10) {
  current_list <- vector("list", 100)
  for (i in 1:100) {
    new_matrix <- CCresultatliste1[[i]][[k]][["consensusMatrix"]]
    current_list[[i]] <- new_matrix
  }
  matrix_liste[[k - 1]] <- current_list
}

# Calculate the mean matrices
mlC <- lapply(matrix_liste, function(matrix_list) {
  Reduce("+", matrix_list) / length(matrix_list)
})

# If you need to explicitly include an empty list as the first element (like meanmatrix1 was), do:
mlC <- c(list(list()), mlC)

save(mlC, file = 'mlC.RData')
```


#Run Consensus clustering
The Consensus cluster results from above is run in a part from CC-package to produce CDF, matrices and cluster numbers based on 100 combined matrices. 
```{r run CC procedure }
#load("sepCCm.rda") #the scaled imputed datasets from MICE 
#load("mlC.rda") #mlC from the CC runs, where the 100 datasets are combined to a meanmatrix for cluster 2-10. 

ml <- mlC
d <- sepCCm[[1]]
res <- list()
maxK <- 10
verbose <- T
finalLinkage <- "average"
plot <- NULL
colorList = list()

colorM <- matrix(ncol = 10)
#colorM = rbind(colorM,colorList[[]])

#18 colors for marking different clusters
thisPal <- c("#A6CEE3","#1F78B4","#B2DF8A","#33A02C","#FB9A99","#E31A1C","#FDBF6F","#FF7F00","#CAB2D6","#6A3D9A","#FFFF99","#B15928",
             "#bd18ea", #magenta
             "#2ef4ca", #aqua
             "#f4cced", #pink,
             "#f4cc03", #lightorange
             "#05188a", #navy,
             "#e5a25a", #light brown
             "#06f106", #bright green
             "#85848f", #med gray
             "#000000", #black
             "#076f25", #dark green
             "#93cd7f",#lime green
             "#4d0776", #dark purple
             "#ffffff" #white
)

myPal = function(n=10){
  #returns n colors
  seq = rev(seq(0,255,by=255/(n)))
  palRGB = cbind(seq,seq,255)
  rgb(palRGB,maxColorValue=255)
}
tmyPal <- myPal(10)

triangle = function(m,mode=1){
  #mode=1 for CDF, vector of lower triangle.
  #mode==3 for full matrix.
  #mode==2 for calcICL; nonredundant half matrix coun
  #mode!=1 for summary 
  n=dim(m)[1]  #n er da dimensjonen pe dataene, for 1 er antall rader; antall pasienter i datasettet
  nm = matrix(0,ncol=n,nrow=n) #lager en matrix med 0 verdier med like mange rader og kolonner som matrix
  fm = m
  
  
  nm[upper.tri(nm)] = m[upper.tri(m)] #set only upper half of the matrix as our matrix
  
  fm = t(nm)+nm
  diag(fm) = diag(m)
  
  nm=fm
  nm[upper.tri(nm)] = NA
  diag(nm) = NA
  vm = m[lower.tri(nm)]
  
  if(mode==1){
    return(vm) #vector 		
  }else if(mode==3){
    return(fm) #return full matrix
  }else if(mode == 2){
    return(nm) #returns lower triangle and no diagonal. no double counts.
  }
  
}


CDF=function(mlC,breaks=100){
  #plot CDF distribution
  plot(c(0),xlim=c(0,1),ylim=c(0,1),col="white",bg="white",xlab="consensus index",ylab="CDF",main="consensus CDF", las=2)
  k=length(mlC)
  this_colors = rainbow(k-1)
  areaK = c()
  for (i in 2:length(mlC)){
    v=triangle(mlC[[i]],mode=1)
    
    #empirical CDF distribution. default number of breaks is 100    
    h = hist(v, plot=FALSE, breaks=seq(0,1,by=1/breaks))
    h$counts = cumsum(h$counts)/sum(h$counts)
    
    #calculate area under CDF curve, by histogram method.
    thisArea=0
    for (bi in 1:(length(h$breaks)-1)){
      thisArea = thisArea + h$counts[bi]*(h$breaks[bi+1]-h$breaks[bi]) #increment by height by width
      bi = bi + 1
    }
    areaK = c(areaK,thisArea)
    lines(h$mids,h$counts,col=this_colors[i-1],lwd=2,type='l')
  }
  legend(0.8,0.5,legend=paste(rep("",k-1),seq(2,k,by=1),sep=""),fill=this_colors)
  
  #plot area under CDF change.
  deltaK=areaK[1] #initial auc at k=2
  for(i in 2:(length(areaK))){
    #proportional increase relative to prior K.
    deltaK = c(deltaK,( areaK[i] - areaK[i-1])/areaK[i-1])
  }
  plot(1+(1:length(deltaK)),y=deltaK,xlab="k",ylab="relative change in area under CDF curve",main="Delta area",type="b")
}
setClusterColors = function(past_ct,ct,colorU,colorList){
  #description: sets common color of clusters between different K
  newColors = c()
  if(length(colorList)==0){
    #k==2
    newColors = colorU[ct]
    colori=2
  }else{
    newColors = rep(NULL,length(ct))
    colori = colorList[[2]]
    mo=table(past_ct,ct)
    m=mo/apply(mo,1,sum)
    for(tci in 1:ncol(m)){ # for each cluster
      maxC = max(m[,tci])
      pci = which(m[,tci] == maxC)				
      if( sum(m[,tci]==maxC)==1 & max(m[pci,])==maxC & sum(m[pci,]==maxC)==1  )  {
        #if new column maximum is unique, same cell is row maximum and is also unique
        ##Note: the greatest of the prior clusters' members are the greatest in a current cluster's members.
        newColors[which(ct==tci)] = unique(colorList[[1]][which(past_ct==pci)]) # one value
      }else{ #add new color.
        colori=colori+1
        newColors[which(ct==tci)] = colorU[colori]
      }
    }
  }
  return(list(newColors,colori,unique(newColors) ))
}
##plot scale
colBreaks=NA
if(is.null(tmyPal)==TRUE){
  colBreaks=10
  tmyPal = myPal(colBreaks)
}else{
  colBreaks=length(tmyPal)
}
sc = cbind(seq(0,1,by=1/( colBreaks) )); rownames(sc) = sc[,1]
sc = cbind(sc,sc)
heatmap(sc, Colv=NA, Rowv=NA, symm=FALSE, scale='none', col=tmyPal, na.rm=TRUE,labRow=rownames(sc),labCol=F,main="consensus matrix legend")

for (tk in 2:maxK){
  if(verbose){
    message(paste("consensus ",tk))
  }
  fm = mlC[[tk]]
  hc=hclust( as.dist( 1 - fm ), method=finalLinkage);
  message("clustered")	
  ct = cutree(hc,tk)
  names(ct) = colnames(d)
  if(any(class(d)=="dist")){
    names(ct) = colnames(as.matrix(d))
  }
  c = fm
?cutree  
  colorList = setClusterColors(res[[tk-1]][[3]],ct,thisPal,colorList)
  pc = c
  pc=pc[hc$order,] #pc is matrix for plotting, same as c but is row-ordered and has names and extra row of zeros.
  
  
  
  if(!is.null(plot) && plot=="pngBMP"){
    pc = pc[,hc$order ] #mod for no tree
    pc = rbind(pc,0)
    #no dendrogram if pngBMP
    oc = colorList[[1]][hc$order] #mod for no tree
    heatmap(pc, Colv = NA, Rowv = NA, symm = FALSE, scale = "none", col = tmyPal, na.rm = TRUE, labRow = F, labCol = F, mar = c(5, 5), main = paste("consensus matrix k=", 
                                                                                                                                                    tk, sep = ""), ColSideCol = oc)
  }else{
    pc = rbind(pc,0)
    #former with tree:
    heatmap(pc, Colv=as.dendrogram(hc), Rowv=NA, symm=FALSE, scale='none', col=tmyPal, na.rm=TRUE,labRow=F,labCol=F,mar=c(5,5),main=paste("consensus matrix k=",tk,sep="") , ColSideCol=colorList[[1]])
  }
  
  legend("topright",legend=unique(ct),fill=unique(colorList[[1]]),horiz=FALSE )
  
  res[[tk]] = list(consensusMatrix=c,consensusTree=hc,consensusClass=ct,mlC=mlC[[tk]],clrs=colorList)
  colorM = rbind(colorM,colorList[[1]]) 
}
CDF(mlC)

#find icl

icl = calcICL(res)

icl[["clusterConsensus"]]
icl[["itemConsensus"]]


```

```{r cluster numbers}
#add column with patientid
for(i in 1:100) {
  sepsisimp[[i]]$id <- sepsis$nnid #list of id-numbers are added 
}

#combine all the imputated datasets and find meanvalues of variables from all the imputations 
ensamb1 <- bind_rows(sepsisimp)

meansepsisvalues <- ensamb1 %>% #making a combined dataset from all the imputed datasets
  group_by(ensamb1$id) %>% 
  summarize_if(is.numeric, mean)

## Backtransform all variables in meansepsisvalues
meansepsisvalues$Creatinine <- exp(meansepsisvalues$logcreatinin)
meansepsisvalues$Lactate <- exp(meansepsisvalues$loglactate)
meansepsisvalues$Crp <- exp(meansepsisvalues$logcrp)
meansepsisvalues$Bun <- exp(meansepsisvalues$logbun)
meansepsisvalues$Alat <- exp(meansepsisvalues$logalat)
meansepsisvalues$Alp <- exp(meansepsisvalues$logalp)
meansepsisvalues$Leucocytes <- exp(meansepsisvalues$logleuc)
meansepsisvalues$Bilirubin <- exp(meansepsisvalues$logbilirubin)
meansepsisvalues$ESR <- exp(meansepsisvalues$logesr)
meansepsisvalues$TroponinT <- exp(meansepsisvalues$logtroponint)
meansepsisvalues$INR <- exp(meansepsisvalues$loginr)
meansepsisvalues$Glucose <- exp(meansepsisvalues$logglu)
meansepsisvalues$Thrombocytes <- exp(meansepsisvalues$logtromb)
meansepsisvalues$paO2 <- exp(meansepsisvalues$logpa02)

meansepsisvalues <- meansepsisvalues[, !names(meansepsisvalues) %in% c("logtroponint","logtromb","logleuc", "logpaO2", "loglactate", "logglu", "logcrp", "Monocytes","logesr", "logalp", "logalat", "loginr", "logcreatinin", "logbilirubin","logalat","logpa02","logalp", "logbun","ensamb1$id")]

#Find medianvalues, IQR of the datasets
k <- 4 #cluster number selected
n <-353 #number of patients"
CClist <- list()
CClist <- res[[k]][["consensusClass"]][1:n]
tableCCprob25 <-aggregate(meansepsisvalues, by=list(CClist), quantile, probs= 0.25)
tableCCprob75 <-aggregate(meansepsisvalues, by=list(CClist), quantile, probs= 0.75)
tableCCmedian <-aggregate(meansepsisvalues, by=list(CClist), median)


```


